---
title: "Chapter 9 Data Aggregation Packages"
author: "Derek Chiu"
date: "Monday, December 08, 2014"
output: slidy_presentation
---

## Introduction

Data aggregation and data visualization go hand in hand. You usually cannot do one without the other.

In R, there are a number of handy packages that allow us to do data aggregation

- `plyr`
- `dplyr`
- `reshape2`
- `tidyr`

## plyr

Package for split-apply-combine notion of data aggregation.

- Data set is __split__ on some variable
- A function is __applied__ on each split
- Results from each split __combined__ together back to one data set

There is a family of functions with `??ply` where the first __?__ can be one of "a", "d", or "l", and the second __?__ can be one of "a", "d", "l", or "_".

- d = data frame
- a = array
- l = list
- _ = discard

The most common function used is `ddply`, taking a `data.frame` and getting back a `data.frame`.

## Example 1: summarize() function

```{r}
library(plyr)
library(gapminder)
gDat <- gapminder
avg_lifeExp <- ddply(gDat, ~ country, summarize, meanLE = round(mean(lifeExp), 2))
```

```{r, echo = FALSE, results = 'asis'}
knitr::kable(avg_lifeExp)
```

## Example 2: custom functions

```{r}
get_meanLE <- function(data) {
	lifeExp <- data$lifeExp
	meanLE <- round(mean(lifeExp), 2)
	setNames(meanLE, "meanLE")
}

avg_lifeExp2 <- ddply(gDat, .(country), get_meanLE)
```

```{r, echo = FALSE, results = 'asis'}
knitr::kable(avg_lifeExp2)
```

## Exercise 1: Country-Specific Linear Regression

Write a function to fit a __linear model__ of life expectancy as a function of year. Apply this function to every country to get country-specific linear fits. Combine all the linear fits into a list, where each element is a fit of a linear model for a different country.

__Hint:__ Write a custom function.

```{r, echo = FALSE}
regLE <- function(data, offset = 1952) {
  lm(lifeExp ~ I(year - 1952), data)
}

all_models <- dlply(gDat, ~ country, regLE)
```

## Example 3: Country-Specific Coefficients

```{r}
all_models[["Canada"]]

coefs <- ldply(all_models, coef)
```

```{r, echo = FALSE, results = 'asis'}
knitr::kable(head(coefs))
```

## dplyr

Essentially a wrapper around `plyr` specifically for `data.frames`. For large data sets, computational time is faster because implementation uses C++. In most cases, the difference is not significant.

Load `plyr` before `dplyr`.

```{r}
library(dplyr)
```

## The forward-pipe operator

`%>%` is known as the `magrittr` forward-pipe operator. It is from the `magrittr` package and very useful in conjunction with `dplyr`. Useful when chaining together multiple commands on data being continuously updated.

Logic is similar to piping in Linux.

Keyboard shortcut is __Ctrl + Shift + M__.

The following lines of code are equivalent

```{r}
dim(head(gDat))

gDat %>%
	head %>%
	dim
```

## tbl_df and glimpse()
A `tbl_df` is a wrapper around the usual `data.frame`. It is advantageous because it only prints the first few rows without specifying `head()`.

```{r}
gDat_tbl <- tbl_df(gDat)
gDat_tbl
```

The `glimpse()` function from `dplyr` is an alternative to our familiar `str()` for looking at the structure of data sets but shows minimal information.

```{r}
glimpse(gDat_tbl)
```

## Essential dplyr operations

- `filter()`: chooses rows with matching conditions (like `subset()`)
- `select()`: chooses columns with matching conditions
- `rename()`: renames columns
- `mutate()`: adds new variables or changes existing ones
- `group_by()` group by one or more variables
- `arrange()`: sorts variables
- `summarize()`: summarises multiple values to a single value

## Example 4: Using the operations

```{r}
hDat <- gDat_tbl %>%
	filter(continent != "Oceania") %>%
	select(-lifeExp) %>%
	rename(nation = country) %>%
	mutate(gdp = pop * gdpPercap) %>%
	group_by(nation) %>%
	summarize(mean_gdp = mean(gdp)) %>%
	arrange(desc(mean_gdp))
```

```{r, echo = FALSE, results = 'asis'}
knitr::kable(head(hDat))
```

`summarize()` is very powerful: it allows you to operate on new variables created in real-time.

## Exercise 2: Change in GDP per capita by Continent
Create a graph that shows how average GDP per capita is changing over time for each continent _except_ Asia.

__Hint:__ For each `continent` and `year` combo, you want a value for the average `gdpPercap`.

_Bonus: Make the ordering of continents consistent between data and legend._

```{r, echo = FALSE}
ex2 <- gDat_tbl %>%
	filter(continent != "Asia") %>%
	group_by(continent, year) %>%
  summarize(mean_gdpPercap = mean(gdpPercap)) %>%
	as.data.frame %>%
	mutate(continent = reorder(continent, mean_gdpPercap))
```

```{r, echo = FALSE}
library(ggplot2)
ggplot(ex2, aes(factor(year), mean_gdpPercap, 
								colour = continent, group = continent))+
  geom_line(lwd = 1)+
  xlab("Year")+
  ylab("Average GDP per capita")+
  ggtitle("Change in Average Life Expectancy by Continent")+
	theme(plot.title = element_text(face = "bold"))+
  scale_colour_manual(values = c(2:5),
                      name = "Continent")+
	guides(color = guide_legend(reverse = TRUE))
```

## reshape2
Changes data between __wide__ and __long__ format.

__Wide format:__

- easy on the eye
- each variable has a different column
- contingency table format

__Long format:__

- harder to analyze
- better for use in making plots
- one column for values
- indicator column that explains what variable the value belongs to
- many repeated entries

## melt and cast

```{r}
library(reshape2)
```

- `melt()` converts data format from wide to long
- `cast()` converts data format from long to wide

We will focus on `dcast()` for `data.frames`.

## Example 5: Reshaping Data 
In our previous exercise, the `ex2` dataset was in long format:

```{r}
head(ex2)
```

It is difficult to analyze time series trends, so we should change it to wide format using `dcast()`:

```{r}
ex2_wide <- dcast(ex2, continent ~ year, value.var = "mean_gdpPercap")
```

```{r, echo = FALSE, results = 'asis'}
knitr::kable(ex2_wide)
```

If we wanted to change it back to long format, we can use `melt()`:

```{r}
ex2_long <- melt(ex2_wide, id.var = "continent", variable.name = "year",
								 value.name = "mean_gdpPercap") %>% 
	arrange(continent, year)
```

```{r, echo = FALSE, results = 'asis'}
knitr::kable(ex2_long)
```

## Exercise 3: Use tidyr

```{r}
library(tidyr)
ex3 <- ex2
```

In the package `tidyr`, there are easier implementations of reshaping data:

- `gather()` is like `melt()`
- `spread()` is like `dcast()`

Read on the function documentations and try to repeat what I did with `ex2` but using functions from `tidyr`. In other words, obtain a `ex3_wide` from `ex3` by using `spread()`, and obtain a `ex3_long` from `ex3_wide` by using `gather()`.

```{r, echo = FALSE}
ex3_wide <- ex3 %>%
	spread(key = year, value = mean_gdpPercap)
```

```{r, echo = FALSE}
ex3_long <- ex3_wide %>%
	gather(key = year, value = mean_gdpPercap, 2:13) %>%
	arrange(continent, year)
```

```{r}
ex3_wide
ex3_long
```

## Other useful packages

- `magrittr`: more operators for piping
- `broom`: converts objects from statistical analysis into tidy data frames

## End

Thank you!

